Assessing US Storm Damage
===========================


###Synopsis: 
This analysis uses publicly available data from the [National Weather Service ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) to answer the following questions:

1. Across the US, which types of weather events are most harmful to public health?
2. Across the US, which types of weather events have the most negative economic impact?


###Data Pre-Processing and Environment Setup

Download the data in an awkwardly zipped format and read it in
```{r Get and read data}
if (!file.exists("StormData.csv.bz2")) 
{
    fileURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
    download.file(fileURL, destfile = 'StormData.csv.bz2')
}
baseData <- read.csv(bzfile('StormData.csv.bz2'), header=TRUE, stringsAsFactors = FALSE)
```

Exploring the baseData dataset
```{r Explore the data}
str(baseData)
names(baseData)
```

Our strategy for determining impact will be as follows:

1. Public health impact ('Harm') will be calculated by adding FATALITIES and INJURIES together
2. Public economic impact ('Cost') will be calculated by adding together Property Damage Expense (PROPDMGEXP) and Crop Damage Expense (CROPDMGEXP)
2.1. It is important to note that the PROPDMGEXP and CROPDMGEXP have been 'deflated' for portability according to the scheme described [here]('https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf'), and to be used it needs to be re-inflated accordingly.

This analysis only uses the following columns from the dataset:

EVTYPE: the type of weather event
FATALITIES: the number of fatalities
INJURIES: the number of injuries
PROPDMG: the amount of property damage in US dollars
PROPDMGEXP: a multiplier for PROPDMG used in inflation/deflation for data portability
CROPDMG: the amount of crop damage in US dollars
CROPDMGEXP: a multiplier for CROPDMG PROPDMG used in inflation/deflation for data portability

Create a second, more compact dataset that contains only the columns we will be working with
```{r Extract columns}
stormData <- data.frame(EventType = baseData$EVTYPE,
                        Fatalities = baseData$FATALITIES,
                        Injuries = baseData$INJURIES,
                        PropDmg = baseData$PROPDMG,
                        pdFactor = baseData$PROPDMGEXP,
                        CropDmg = baseData$CROPDMG,
                        cdFactor = baseData$CROPDMGEXP,
                        stringsAsFactors = FALSE)
```

Now keep only those rows that have a death OR an injury OR a property damage cost OR a crop damagae cost.
```{r Remove empty rows}
stormData <- stormData[stormData$Fatalities > 0 | stormData$Injuries > 0 | 
                           stormData$PropDmg > 0 | stormData$CropDmg > 0, ]
```

Next we need to 're-inflate' the property and crop damage fields. In a scheme that clearly dates to the days when 32kb cost a lot of money, the database saves a (ridiculously) marginal amount of space by using the pdFactor and cdFactor fields to represent a multiple: 

* H/h is the multiple for 100
* K/k is the multiple for 1,000
* M/m is the multiple for 1,000,000
* B/b is the multiple for 1,000,000

So, for example, if in the CropDmg field we have 9.8, and in the cdFactor field we have M, then we know the crop damage was $9,800,000.

```{r Inflate Damages}
for (i in 1:nrow(stormData)){
    if (tolower(stormData$pdFactor[i]) == 'h') {
        stormData$PropDmg[i] <- stormData$PropDmg[i] * 100
    }
    else if (tolower(stormData$pdFactor[i]) == 'k') {
        stormData$PROPDMG[i] <- stormData$PROPDMG[i] * 1000
    }
    else if (tolower(stormData$pdFactor[i]) == 'm') {
        stormData$PropDmg[i] <- stormData$PropDmg[i] * 1000000
    }
    else if (tolower(stormData$pdFactor[i]) == 'b') {
        stormData$PropDmg[i] <- stormData$PropDmg[i] * 1000000000
    }
    
    if (tolower(stormData$cdFactor[i]) == 'h') {
        stormData$CropDmg[i] <- stormData$CropDmg[i] * 100
    }
    else if (tolower(stormData$cdFactor[i]) == 'k') {
        stormData$CropDmg[i] <- stormData$CropDmg[i] * 1000
    }
    else if (tolower(stormData$cdFactor[i]) == 'm') {
        stormData$CropDmg[i] <- stormData$CropDmg[i] * 1000000
    }
    else if (tolower(stormData$cdFactor[i]) == 'b') {
        stormData$CropDmg[i] <- stormData$CropDmg[i] * 1000000000
    }
}
```
Now we calculate totals by event type and sort (decreasing)
```{r Sum and Sort}
totalFatalities <- aggregate(stormData$Fatalities, by = list(EventType = stormData$EventType), FUN = sum)
totalInjuries <- aggregate(stormData$Injuries, by = list(EventType = stormData$EventType), FUN = sum)
totalPropDmg <- aggregate(stormData$PropDmg, by = list(EventType = stormData$EventType), FUN = sum)
totalCropDmg <- aggregate(stormData$CropDmg, by = list(EventType = stormData$EventType), FUN = sum)

colnames(totalFatalities)[2] <- 'Total'
colnames(totalInjuries)[2] <- 'Total'
colnames(totalPropDmg)[2] <- 'Total'
colnames(totalCropDmg[2]) <- 'Total'

totalFatalities <- totalFatalities[order(totalFatalities$Total, decreasing = T), ]
totalInjuries <- totalInjuries[order(totalInjuries$Total, decreasing = T), ]
totalPropDmg <- totalPropDmg[order(totalPropDmg$Total, decreasing = T), ]
totalCropDmg <- totalCropDmg[order(totalCropDmg$Total, decreasing = T), ]
```

For the sake of chart readability, we are going to limit our analysis to the top 12 event types

```{r Top Limiter}
totalFatalities <- totalFatalities[1:12, ]
totalInjuries <- totalInjuries[1:12, ]
totalPropDmg <- totalPropDmg[1:12, ]
totalCropDmg <- totalCropDmg[1:12, ]
```

















